---
title: "COVID-19 Portugal data"
output: 
  github_document:
    dev: svg
---

```{r, eval=FALSE, include=FALSE}
rmarkdown::render('_README.Rmd', output_file = 'README.md')
```

```{r, eval=FALSE, include=FALSE}
rmarkdown::clean_site()
rmarkdown::render_site(quiet = TRUE)
```


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all()
library(pdftools)
library(rvest)
library(dplyr)
library(tibble)
library(DT)
library(anytime)
library(glue)
library(reshape2)
library(zoo)
library(xml2)
library(usethis)
library(ggplot2)
library(ggrepel)
library(plotly)

Sys.setlocale('LC_TIME', 'en_GB.UTF-8')
```

> R package with latest data scrapped from official sources

It downloads the [daily report](https://covid19.min-saude.pt/relatorio-de-situacao/) from DGS and stores this in data-friendly format under `/data` directory.

A more detailed analysis of this data is [available here](https://averissimo.github.io/covid19-analysis/portugal.html)

# Check for new reports

```{r, eval=FALSE}
download.updated.pt()
```

```{r, message=FALSE, include=FALSE}
dat <- download.updated.pt()

dgs.pt.new     <- dat$dgs.pt 
covid19.pt.new <- dat$cdc.eu

dgs.pt <- tibble()
tryCatch(dgs.pt <- covid19.pt.data::dgs.pt, error = function(err) { })

# DGS PT
if (nrow(dgs.pt.new) != nrow(dgs.pt)) {
  dgs.pt <- dgs.pt.new 
  usethis::use_data(dgs.pt, overwrite = TRUE)
  readr::write_csv(dgs.pt, path = '../data/dgs_pt.csv')
}

# EU CDC
covid19.pt <- tibble()
tryCatch(covid19.pt <- covid19.pt.data::covid19.pt, error = function(err) { })

if (nrow(covid19.pt.new) != nrow(covid19.pt)) {
  covid19.pt <- covid19.pt.new
  usethis::use_data(covid19.pt, overwrite = TRUE)
  readr::write_csv(covid19.pt, path = '../data/covid19_pt.csv')
}
```

```{r, echo=FALSE}
dgs.pt <- tibble()
tryCatch(dgs.pt <- covid19.pt.data::dgs.pt, error = function(err) { })
```

```{r, echo=FALSE}
age.data <- dgs.pt %>% 
  reshape2::melt(id.vars = c('country', 'date'), variable.name = 'type') %>% 
  group_by(country, type) %>% 
  filter(grepl('(confirmed_|death_)', type)) %>% 
  arrange(date) %>% 
  ungroup() %>% 
  mutate(gender = if_else(grepl('.*_m_*', type), 'men', 'women'),
         age_type = if_else(grepl('confirmed_', type), 'confirmed', 'death'),
         type = gsub('.*_(m|w)_', '', type),
         value = if_else(gender == 'men', value * -1, value %>% as.double))
```

## Data for Portugal

```{r, echo=FALSE, warning=FALSE, fig.width=10, fig.height=7}
dgs.pt %>% 
  reshape2::melt(id.vars = c('country', 'date'), variable.name = 'type', measure.vars = c('tests', 'deaths', 'confirmed', 'hospitalized', 'in.icu', 'recovered')) %>% 
  mutate(type = 
           gsub('confirmed', 'Confirmed cases', type) %>% 
           gsub('deaths', 'Deaths', .) %>% 
           gsub('hospitalized', 'Hospitalized', .) %>% 
           gsub('recovered', 'Recovered', .) %>% 
           gsub('tests', 'Tests', .) %>% 
           gsub('in.icu', 'In ICU', .)) %>% 
  group_by(country, type) %>% 
  arrange(date) %>% 
  mutate(label = if_else(date == last(date), 
                         paste0(type, ': ', 
                                format(value, digits = 2, big.mark =',', trim = FALSE)), 
                         NA_character_)) %>% 
  ggplot(aes(x = date, y = value, color = type, label = label, fill = type)) +
    geom_line(size = 2) + 
    geom_point(size = 2.5) +
    ggrepel::geom_label_repel(color = 'white', 
                              segment.color = 'black', 
                              na.rm = TRUE, 
                              min.segment.length = 0, 
                              nudge_x = -2,
                              nudge_y = -.5,
                              segment.alpha = .8,
                              size = 3.5,
                              seed = 1985) +
    labs(x = 'Date', 
         y = 'Number of individuals',
         title = 'Individuals that were Infected, Recovered and Died',
         subtitle = 'Showing in Logarithmic scale (base 10)') +
    scale_y_continuous(trans = 'log10', labels = function(ix) { format(ix, big.mark = ',', scientific = FALSE) }) +
    scale_color_viridis_d(end = .7) +
    scale_fill_viridis_d(end = .7) +
    theme_minimal() +
    theme(legend.position = 'none')
```

```{r, echo=FALSE, warning=FALSE, fig.width=10, fig.height=7}
plot.data <- dgs.pt %>% 
  reshape2::melt(id.vars = c('country', 'date'), variable.name = 'type', measure.vars = c('tests', 'deaths', 'confirmed', 'hospitalized', 'in.icu', 'recovered')) %>% 
  mutate(type = 
           gsub('confirmed', 'Confirmed cases', type) %>% 
           gsub('deaths', 'Deaths', .) %>% 
           gsub('hospitalized', 'Hospitalized', .) %>% 
           gsub('recovered', 'Recovered', .) %>% 
           gsub('tests', 'Tests', .) %>% 
           gsub('in.icu', 'In ICU', .)) %>% 
  group_by(country, type) %>% 
  arrange(desc(date)) %>% 
  mutate(value = zoo::rollapply(value, 
                                2, 
                                function(ix) { if(length(ix) <= 1) { return(ix) } else { ix[1] - sum(ix[-1]) } }, fill = c(0, 0, 0), align = 'left', partial = TRUE)) %>%
  group_by(country, type) %>% 
  arrange(date) %>% 
  filter(date > min(date)) %>%  ## values will not be for that day
  build.labels('value', 'type') 

roundUp <- function(x) 10^ceiling(log10(x))
my.breaks <- c(-100, -40, -20, -10, -5, -2, 0, 2, 5, 10, 30, 100, 300, 1000, 3000, 10000, 20000)

my.breaks <- c(my.breaks, max(plot.data$value, na.rm = TRUE) %>% roundUp(), -1 * (abs(min(plot.data$value, na.rm = TRUE)) %>% roundUp())) %>% 
  unique %>% 
  sort

plot.data %>% 
  ggplot(aes(x = date, y = value, color = type, label = label, fill = type)) +
    geom_line(size = 2) + 
    geom_point(size = 2.5) +
    ggrepel::geom_label_repel(color = 'white', 
                              segment.color = 'black', 
                              na.rm = TRUE, 
                              min.segment.length = 0, 
                              nudge_x = -2,
                              nudge_y = -.5,
                              segment.alpha = .8,
                              size = 3.5,
                              seed = 1985) +
    labs(x = 'Date', 
         y = 'Number of individuals',
         title = 'Individuals that were Infected, Recovered and Died',
         subtitle = 'Showing in a (pseudo) logarithmic scale') +
    scale_color_viridis_d(end = .7) +
    scale_fill_viridis_d(end = .7) +
    scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10), 
                       labels = function(ix) { format(ix, big.mark = ',', scientific = FALSE) },
                       breaks =  my.breaks) +
    theme_minimal() +
    theme(legend.position = 'none')
```


# New cases / deaths by age groups

Only showing 1 day *(`r format(age.data$date %>% max, "%B %d")`)*

```{r, echo=FALSE,fig.width=4}
age.data.all.new <- age.data %>% 
  group_by(country, type, gender, age_type) %>% 
  arrange(desc(date)) %>% 
  mutate(value = zoo::rollapply(value, 2, function(ix) { if(length(ix) <= 1) { return(ix) } else { ix[1] - sum(ix[-1]) } }, fill = c(0, 0, 0), align = 'left', partial = TRUE)) %>%
  filter(value != 0) %>% 
  filter(date == max(date) & value != 0) %>% 
  ungroup() %>% 
  select(type, gender, age_type, value) %>% 
  group_by(type, gender, age_type) %>% 
  reshape2::dcast(type + gender ~ age_type, mean, fill = 0) %>% 
  mutate(label.confirmed = if_else(confirmed != 0, format(abs(confirmed), big.mark = ',', trim = TRUE), NA_character_),
         label.death = if_else(death != 0, 
                               if_else(is.na(confirmed) | confirmed == 0, format(abs(death), big.mark = ',', trim = TRUE),
                                       paste0(format(abs(death), big.mark = ',', trim = TRUE))),
                               NA_character_))

confirmed.max <- age.data.all.new %>%  pull(confirmed) %>% max
death.max <- age.data.all.new %>% pull(death) %>% max

print(age.data.all.new %>%
  ggplot(aes(x = confirmed, y = type, fill = gender)) +
  geom_bar(stat = 'identity') + 
  ggrepel::geom_label_repel(aes(label = label.confirmed, fill = gender), color = 'white', direction = 'x', seed = 1985, size = 3.5,
                        nudge_x = ifelse(age.data.all.new$gender == 'men', -1, 1),
                        show.legend = FALSE,
                        na.rm = TRUE) +
  expand_limits(x =c(-1 * confirmed.max, confirmed.max)) +
  scale_x_continuous('', labels = function(ix) { return(abs(ix)) }) +
  scale_y_discrete(limits = age.data$type %>% unique %>% sort) + 
  scale_fill_viridis_d('', end = .8) + 
  labs(title = "Confirmed cases",
       subtitle = 'Only for single day ({format(age.data$date %>% max, "%B %d")})' %>% glue,
       y = 'Age group',
       x = 'Confirmed Cases') +
  theme_minimal() + 
  theme(legend.position = 'bottom'))
print(age.data.all.new %>%
  ggplot(aes(x = death, y = type, fill = gender)) +
  geom_bar(stat = 'identity') + 
  ggrepel::geom_label_repel(aes(label = label.death, fill = gender), color = 'white', direction = 'x', seed = 1985, size = 3.5,
                            nudge_x = ifelse(age.data.all.new %>% filter(!is.na(label.death)) %>% pull(gender) == 'men', -1, 1),
                            show.legend = FALSE,
                            na.rm = TRUE) +
  expand_limits(x =c(-1 * death.max, death.max)) +
  scale_x_continuous('', labels = function(ix) { return(abs(ix)) }) +
  scale_fill_viridis_d('', end = .8) + 
  labs(title = "Total number of Deaths by age group",
       subtitle = 'Only for single day ({format(age.data$date %>% max, "%B %d")})' %>% glue,
       y = 'Age group',
       x = 'Deaths') +
  theme_minimal() + 
  theme(legend.position = 'bottom'))
```


# Cases / Deaths by age groups {.tabset .tabset-fade .tabset-pills}

```{r, result='asis', echo=FALSE,fig.width=4}
age.data.all <- age.data %>% 
  filter(date == max(date) & value != 0) %>% 
  select(type, gender, age_type, value) %>% 
  reshape2::dcast(type + gender ~ age_type, mean, fill = 0) %>% 
  mutate(label.confirmed = if_else(confirmed != 0, format(abs(confirmed), big.mark = ',', trim = TRUE), NA_character_),
         label.death = if_else(death != 0, 
                               paste0(format(abs(death), big.mark = ',', trim = TRUE),
                                      ' (',
                                      scales::percent(abs(death/confirmed), accuracy = 0.01, big.mark = ','),
                                      ')'), 
                               NA_character_))

confirmed.max <- age.data.all %>%  pull(confirmed) %>% max
death.max <- age.data.all %>% pull(death) %>% max

print(age.data.all %>%
  ggplot(aes(x = confirmed, y = type, fill = gender)) +
  geom_bar(stat = 'identity') + 
  ggrepel::geom_label_repel(aes(label = label.confirmed, fill = gender), color = 'white', direction = 'x', seed = 1985, size = 3.5,
                        nudge_x = ifelse(age.data.all$gender == 'men', -1, 1),
                        show.legend = FALSE,
                        na.rm = TRUE) +
  expand_limits(x =c(-1 * confirmed.max, confirmed.max)) +
  scale_x_continuous('', labels = function(ix) { return(abs(ix)) }) +
  scale_y_discrete(limits = age.data$type %>% unique %>% sort) + 
  scale_fill_viridis_d('', end = .8) + 
  labs(title = "Confirmed cases",
       y = 'Age group',
       subtitle = '',
       caption = '',
       x = 'Confirmed Cases') +
  theme_minimal() + 
  theme(legend.position = 'bottom'))
print(age.data.all %>%
  ggplot(aes(x = death, y = type, fill = gender)) +
  geom_bar(stat = 'identity') + 
  ggrepel::geom_label_repel(aes(label = label.death, fill = gender), color = 'white', direction = 'x', seed = 1985, size = 3.5,
                            nudge_x = ifelse(age.data.all %>% filter(!is.na(label.death)) %>% pull(gender) == 'men', -1, 1),
                            show.legend = FALSE,
                            na.rm = TRUE) +
  expand_limits(x =c(-1 * death.max, death.max)) +
  scale_x_continuous('', labels = function(ix) { return(abs(ix)) }) +
  scale_fill_viridis_d('', end = .8) + 
  labs(title = "Total number of Deaths by age group",
       subtitle = 'Percentage shows the mortality rate',
       caption = 'Mortality rate = \'deaths\' / \'confirmed cases\'',
       y = 'Age group',
       x = 'Deaths') +
  theme_minimal() + 
  theme(legend.position = 'bottom'))
```

# Data {.tabset .tabset-fade .tabset-pills}

## Data from DGS

Only showing last 10 days

```{r, echo=FALSE}
dgs.pt %>% arrange(desc(date)) %>% top_n(10, date) %>% knitr::kable()
```

## Data from EU CDC updated

Only showing last 10 days.

Don't be alarmed with the first line being in the future, EU CDC date always refer to the situation on the previous day.

```{r, message=FALSE, echo=FALSE}
covid19.pt %>% head(10) %>% knitr::kable()
```




