---
title: "COVID-19 Portugal data"
output: 
  github_document:
    dev: svg
---

```{r, eval=FALSE, include=FALSE}
rmarkdown::render('_README.Rmd', output_file = 'README.md')
```

```{r, eval=FALSE, include=FALSE}
rmarkdown::clean_site()
rmarkdown::render_site(quiet = TRUE)
```

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
devtools::load_all()
library(pdftools)
library(rvest)
library(dplyr)
library(tibble)
library(DT)
library(anytime)
library(glue)
library(reshape2)
library(zoo)
library(xml2)
library(ggplot2)
library(ggrepel)
library(plotly)

Sys.setlocale('LC_TIME', 'en_GB.UTF-8')
```

```{r, message=FALSE, include=FALSE}
dat <- download.updated.pt()

dgs.pt.new     <- dat$dgs.pt 

dgs.pt <- tibble()
tryCatch(dgs.pt <- covid19.pt.data::dgs.pt, error = function(err) { })

# DGS PT
if (nrow(dgs.pt.new) != nrow(dgs.pt) || any(is.na(dgs.pt.new[1,]) != is.na(dgs.pt[1,]), na.rm = FALSE)) {
  dgs.pt <- dgs.pt.new 
  usethis::use_data(dgs.pt, overwrite = TRUE)
  readr::write_csv(dgs.pt, file = '../data/dgs_pt.csv')
}

```

> R package with latest data scrapped from official sources *(last data from `r format(max(dgs.pt.new$date), "%A, %B %d, %Y")`)*

It downloads the [daily report](https://covid19.min-saude.pt/relatorio-de-situacao/) from DGS and stores this in data-friendly format under `/data` directory. It also retrieves age data from the DGS' [ESRI dashboard](https://covid19.min-saude.pt/ponto-de-situacao-atual-em-portugal/)

If you are here just for the data, this is what you want: 

* [dgs_pt.csv](raw/master/data/dgs_pt.csv) *(raw data from Portugal's DGS)*

note: The EU CDC data has been removed as the daily reports from EU have been discontinued in favor of weekly counts.

A mortality analysis of Portugal is [available here](https://averissimo.github.io/covid19-analysis/mortality.html)

# Check for new reports

```{r, eval=FALSE}
download.updated.pt()
```

```{r, echo=FALSE}
dgs.pt <- dgs.pt.new 
```

```{r, echo=FALSE}
age.data <- get.age.data(dgs.pt)
```

## Data for Portugal

Last 5 days _(how each column is changing over these days)_

```{r, echo=FALSE}
dgs.pt %>% arrange(desc(date)) %>% top_n(6, date) %>%
  dplyr::mutate_if(is.integer, function(my.input) {return(zoo::rollapply(data = my.input, width = 2, FUN = function(a) {return(a[1] - mean(a[-1]))}, align = 'left', partial = TRUE))}) %>% 
  dplyr::filter(!is.nan(confirmed)) %>% 
  select(-tests) %>% 
  knitr::kable()
```



```{r totals, echo=FALSE, warning=FALSE, fig.width=10, fig.height=7}
dgs.pt %>% 
  reshape2::melt(id.vars = c('country', 'date'), variable.name = 'type', measure.vars = c('deaths', 'confirmed', 'hospitalized', 'in.icu', 'recovered')) %>% 
  mutate(type = 
           gsub('confirmed', 'Confirmed cases', type) %>% 
           gsub('deaths', 'Deaths', .) %>% 
           gsub('hospitalized', 'Hospitalized', .) %>% 
           gsub('recovered', 'Recovered', .) %>% 
           gsub('tests', 'Suspected individuals', .) %>% 
           gsub('in.icu', 'In ICU', .)) %>% 
  group_by(country, type) %>% 
  arrange(date) %>% 
  mutate(label = if_else(date == last(date), 
                         paste0(type, ': ', 
                                format(value, digits = 2, big.mark =',', trim = FALSE)), 
                         NA_character_)) %>% 
  ggplot(aes(x = date, y = value, color = type, label = label, fill = type)) +
    geom_line(size = 1, alpha = .6) + 
    geom_point(size = 1) +
    ggrepel::geom_label_repel(color = 'white', 
                              segment.color = 'red', 
                              na.rm = TRUE, 
                              min.segment.length = 0, 
                              #nudge_x = -2,
                              #nudge_y = -.5,
                              segment.alpha = .8,
                              max.time = 5,
                              alpha = .85,
                              max.iter = 100000,
                              force = 40,
                              force_pull = 0,
                              max.overlaps = 3,
                              size = 3,
                              seed = 1985) +
    scale_x_date(date_labels = '%B %d (%a)', date_breaks = "5 day", date_minor_breaks = "1 day") +
    labs(x = 'Date', 
         y = 'Number of individuals',
         title = 'Individuals that were Infected, Recovered and Died',
         subtitle = 'Showing in Logarithmic scale (base 10)',
         caption = "{format(max(dgs.pt.new$date), '%A, %B %d, %Y')}\n Label points to the maximum point in time" %>% glue::glue()) +
    scale_y_continuous(trans = 'log10', labels = function(ix) { format(ix, big.mark = ',', scientific = FALSE) }) +
    scale_color_viridis_d(end = .7) +
    scale_fill_viridis_d(end = .7) +
    theme_minimal() +
    theme(legend.position = 'none', axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r differential, echo=FALSE, warning=FALSE, fig.width=10, fig.height=7}
plot.data <- dgs.pt %>% 
  reshape2::melt(id.vars = c('country', 'date'), variable.name = 'type', measure.vars = c('deaths', 'confirmed', 'hospitalized', 'in.icu', 'recovered')) %>% 
  mutate(type = 
           gsub('confirmed', 'Confirmed cases', type) %>% 
           gsub('deaths', 'Deaths', .) %>% 
           gsub('hospitalized', 'Hospitalized', .) %>% 
           gsub('recovered', 'Recovered', .) %>% 
           gsub('tests', 'Suspected individuals', .) %>% 
           gsub('in.icu', 'In ICU', .)) %>% 
  group_by(country, type) %>% 
  arrange(desc(date)) %>% 
  mutate(value = zoo::rollapply(value, 
                                2, 
                                function(ix) { if(length(ix) <= 1) { return(ix) } else { ix[1] - sum(ix[-1]) } }, fill = c(0, 0, 0), align = 'left', partial = TRUE)) %>%
  group_by(country, type) %>% 
  arrange(date) %>% 
  filter(date > min(date)) %>%  ## values will not be for that day
  build.labels('value', 'type', show_plus = TRUE) 

roundUp <- function(x) 10^ceiling(log10(x))
my.breaks <- c(-100, -40, -20, -10, -5, -2, 0, 2, 5, 10, 30, 100, 300, 1000, 3000, 10000, 20000)

my.breaks <- c(my.breaks, max(plot.data$value, na.rm = TRUE) %>% roundUp(), -1 * (abs(min(plot.data$value, na.rm = TRUE)) %>% roundUp())) %>% 
  unique %>% 
  sort

plot.data %>% 
  ggplot(aes(x = date, y = value, color = type, label = label, fill = type)) +
    geom_hline(aes(yintercept = 0), size = .7, alpha = .3, linetype = 3) +
    geom_line(size = 1, alpha = .6) + 
    geom_point(size = 1) +
    ggrepel::geom_label_repel(color = 'white', 
                              segment.color = 'red', 
                              na.rm = TRUE, 
                              min.segment.length = 0, 
                              #nudge_x = -2,
                              #nudge_y = -.5,
                              segment.alpha = .8,
                              max.time = 5,
                              alpha = .85,
                              max.iter = 100000,
                              force = 40,
                              force_pull = 0,
                              size = 3,
                              seed = 1985) +
    labs(x = 'Date', 
         y = 'Number of individuals',
         title = 'Individuals that were Infected, Recovered and Died',
         subtitle = 'Showing in a (pseudo) logarithmic scale',
         caption = "data from {format(max(dgs.pt.new$date), '%A, %B %d, %Y')}\n Label points to the maximum point in time" %>% glue::glue()) +
    scale_x_date(date_labels = '%B %d (%a)', date_breaks = "3 day", date_minor_breaks = "1 day") +
    scale_color_viridis_d(end = .7) +
    scale_fill_viridis_d(end = .7) +
    scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10), 
                       labels = function(ix) { format(ix, big.mark = ',', scientific = FALSE) },
                       breaks =  my.breaks) +
    theme_minimal() +
    theme(legend.position = 'none', axis.text.x = element_text(angle = 45, hjust = 1, size = 5))
```

```{r differential_7days, echo=FALSE, warning=FALSE, fig.width=10, fig.height=7}
plot.data <- dgs.pt %>% 
  reshape2::melt(id.vars = c('country', 'date'), variable.name = 'type', measure.vars = c('deaths', 'confirmed', 'hospitalized', 'in.icu', 'recovered')) %>% 
  mutate(type = 
           gsub('confirmed', 'Confirmed cases', type) %>% 
           gsub('deaths', 'Deaths', .) %>% 
           gsub('hospitalized', 'Hospitalized', .) %>% 
           gsub('recovered', 'Recovered', .) %>% 
           gsub('tests', 'Suspected individuals', .) %>% 
           gsub('in.icu', 'In ICU', .)) %>% 
  group_by(country, type) %>% 
  arrange(desc(date)) %>% 
  mutate(value = zoo::rollapply(value, 
                                2, 
                                function(ix) { if(length(ix) <= 1) { return(ix) } else { ix[1] - sum(ix[-1]) } }, fill = c(0, 0, 0), align = 'left', partial = TRUE)) %>%
  mutate(value = zoo::rollapply(value, 7, mean, align = 'left', fill = c(0,0,0), partial = TRUE)) %>%
  group_by(country, type) %>% 
  arrange(date) %>% 
  filter(date > min(date)) %>%  ## values will not be for that day
  build.labels('value', 'type', digits = 0, show_plus = TRUE) 

roundUp <- function(x) 10^ceiling(log10(x))
my.breaks <- c(-100, -40, -20, -10, -5, -2, 0, 2, 5, 10, 30, 100, 300, 1000, 3000, 10000, 20000)

my.breaks <- c(my.breaks, max(plot.data$value, na.rm = TRUE) %>% roundUp(), -1 * (abs(min(plot.data$value, na.rm = TRUE)) %>% roundUp())) %>% 
  unique %>% 
  sort

plot.data %>% 
  ggplot(aes(x = date, y = value, color = type, label = label, fill = type)) +
    geom_hline(aes(yintercept = 0), size = .7, alpha = .3, linetype = 3) +
    geom_line(size = 1, alpha = .6) + 
    geom_point(size = 1) +
    ggrepel::geom_label_repel(color = 'white',
                              segment.color = 'red', 
                              na.rm = TRUE, 
                              min.segment.length = 0, 
                              #nudge_x = -2,
                              #nudge_y = -.5,
                              segment.alpha = .8,
                              size = 3,
                              alpha = .85,
                              max.time = 5,
                              max.iter = 1000000,
                              max.overlaps = 5,
                              force = 40,
                              force_pull = 0,
                              seed = 1985) +
    labs(x = 'Date', 
         y = 'Number of individuals',
         title = '7-day average of Individuals that were Infected, Recovered and Died',
         subtitle = 'Showing in a (pseudo) logarithmic scale',
         caption = "{format(max(dgs.pt.new$date), '%A, %B %d, %Y')}\n Label points to the maximum point in time" %>% glue::glue()) +
    scale_x_date(date_labels = '%B %d (%a)', date_breaks = "3 day", date_minor_breaks = "1 day") +
    scale_color_viridis_d(end = .7) +
    scale_fill_viridis_d(end = .7) +
    scale_y_continuous(trans = scales::pseudo_log_trans(sigma = 1, base = 10), 
                       labels = function(ix) { format(ix, big.mark = ',', scientific = FALSE) },
                       breaks =  my.breaks) +
    theme_minimal() +
    theme(legend.position = 'none', axis.text.x = element_text(angle = 45, hjust = 1, size = 5))
```

# New cases / deaths by age groups

```{r, echo=FALSE}
age.ix.f <- dgs.pt %>% filter(!is.na(`confirmed_m_00-09`)) %>% pull(date) %>% max
age.data.all <- get.age.data.with.labels(age.data, age.ix.f)
age.data.all.new <- get.age.new.data(age.data, age.data.all, age.ix.f)
```

```{r, results='asis', echo=FALSE}
if(age.ix.f != dgs.pt %>% pull(date) %>% max) {
  cat('*note: data update may be delayed by a couple of hours (since age data was been removed from daily report and we\'ve been retrieving it from DGS\'s covid19 dashboard, which may have a delay)*')
}
```

Age data may show different numbers from absolutes in the daily report. This is out of our control and dependent on the data source _(DGS covid19 dashboard)_.

```{r new_cases_deaths, echo=FALSE,fig.width=4.1}
if(!is.null(age.data.all.new)) {
 my.plots.new <- get.plot.for.new(age.data.all.new, age.ix.f)
 print(my.plots.new$confirmed + theme(plot.title = element_text(size=10), plot.subtitle = element_text(size=8)))
 print(my.plots.new$deaths + theme(plot.title = element_text(size=10), plot.subtitle = element_text(size=8)))
} else {
 cat('Missing age grouping information for this day:', format(age.ix.f)) 
}
```

# All cases / deaths by age groups

Age data may show different numbers from absolutes in the daily report. This is out of our control and dependent on the data source _(DGS covid19 dashboard)_.

```{r total_cases_deaths, echo=FALSE,fig.width=4.1}
if(!is.null(age.data.all.new)) {
 my.plots.all <- get.plot.for.all(age.data.all, age.ix.f, dgs.pt = dgs.pt)
 print(my.plots.all$confirmed + theme(plot.title = element_text(size=9), plot.subtitle = element_text(size=8)))
 print(my.plots.all$deaths + theme(plot.title = element_text(size=9), plot.subtitle = element_text(size=8)))
} else {
 cat('Missing age grouping information for this day:', format(age.ix.f)) 
}
```

# Data {.tabset .tabset-fade .tabset-pills}

## Data from DGS

Only showing last 20 days

```{r, echo=FALSE}
dgs.pt %>% arrange(desc(date)) %>% top_n(20, date) %>% knitr::kable()
```



